# Genetic Groups

## Objective: Assign individuals to genetic groups

Previous research has suggested that there are four different genetic clusters in the gulf: STX, nGOM, nFL (northern Florida), and genFL (general Florida).

The objective of this section is to use the genotype data from the 600K SNP array to assign oysters to genetic groups. 

```{r, include = FALSE}
#library(SNPolisher)
library(tidyverse)
library(hierfstat)
library(adegenet)
library(stringr)
library(ggrepel)
library(pantomime)
library(pegas)
library(patchwork)
library("lattice")
library(maps)
library(mapdata)
library(stringr)
library(data.table)
library(here)
```

```{r, include = FALSE}
#load in functions for PCA and manipulating genind objects
source("~/PCA.R")
source("~/genind.R")
```

## Data Subsetting

Read in SNP data -- combined_filtered data set

* row names are individuals

* columns are loci

* cells are genotypes (11,12,21,22)

```{r}
combined_filtered <-read.table("data/derived/combined_filtered.txt",header=TRUE)
```


Read in Individual data -- sample_info_full data set

* first column is individuals

* 52 other columns of descriptive information on individuals

```{r}
sample_info_full <-read.table("data/derived/sample_info_full.txt",header=TRUE)
```


Produce list of individuals that are from selected lines and take the individuals from the selected lines out of the SNP data set

```{r}
#Produce a list of individuals that are from selected lines (O'Boy or Auburn selected lines)
selected <- sample_info_full %>%
  filter(Collection_data %in% c("Auburn selected line 1", "Auburn selected line 2", "O'Boy selected line")) %>%
    pull(Sample)

#Take the individuals from selected lines out of the SNP data set
genotypes <- combined_filtered %>%
  filter(!row.names(.) %in% selected)
```

Subset the SNP data set by selecting a random 20K SNPs

* should speed up analysis and be just as effective for genetic group assignment

```{r}
set.seed(54)

combined_filtered_20K <- genotypes %>%
  select(sample(seq_len(ncol(.)), size = 20000,replace=FALSE))
```

## Kmeans - All Wild Individuals

Prior groups need to be defined for DAPC

* identify groups initially using k-means clustering

* k-means is a clustering algorithm which finds a given number of groups that maximizes variation among groups

* k-means is run sequentially with increasing number of ks, and solutions can be compared using BIC

* ideally, lowest BIC is best clustering solution 

* k-means can be done much faster on the PCA data than the raw data

* transform the data using PCA with the find.clusters function, asks the user to specify the number of retained PCS interactively

First, Convert SNP data to a genind object

```{r}
#convert SNP data to genind
gen1 <- df2genind(combined_filtered_20K, ncode = 1, sep = "", NA.char = NA, ploidy = 2)
gen1
```

Use find.clusters to identify groups via k-means clustering

```{r}
grp <- find.clusters(gen1, n.pca = 500, choose = FALSE, max.n.clust = 5) # retain 500 PCs, retain 3 clusters

plot(grp$Kstat, type="o", xlab="number of clusters (K)", ylab="BIC",
col="blue", main="Detection based on BIC")
```

Looks like there are three or four groups

Choose three clusters

```{r}
grp <- find.clusters(gen1, n.pca = 500, n.clust = 3)
grp$size
```

Use the k-means clustering and the sample metadata to create a tibble that contains the 380 individuals that are wild collected and 54 columns of descriptive information, including the kmeans group assignment "k_group"

```{r}
# Returns data frame of individuals and their k-means assignment
x <- as_tibble(grp$grp,rownames=NA) %>%
  rownames_to_column(.,var="Sample") %>%
  rename(k_group = value)
```

Make a data set that only has wild individuals

```{r}
sample_info_wild <- sample_info_full %>%
  filter(! Collection_data %in% c("Auburn selected line 1", "Auburn selected line 2", "O'Boy selected line"))
```

Add a column of k_means group to data on individuals

```{r}
sample_data <- sample_info_wild %>%
  left_join(., x, 
              by = "Sample")
```

## PCA - K groups

Use PCAs to confirm there are 3 groups as identified by k means

```{r}
#identify each individual to each k group in the genind
kg<-sample_data %>%
  select(Sample,k_group)

# Assign strata to genind
strata(gen1) <- kg[match(indNames(gen1),sample_data$Sample),]

# set pop to GG/STX
setPop(gen1) <- ~k_group
```

```{r, message=FALSE, warning=FALSE}
#Check PCA eigenvalues
x<- scaleGen(gen1, NA.method = "mean")

# Plot a PCA 
pca <- dudi.pca(x,cent=FALSE,scale=FALSE,scannf=FALSE,nf=4)

#plot1
barplot(pca$eig[1:50],main="All_Samples_PCA_eigenvalues", col=heat.colors(50))
```

```{r message = FALSE,warning = FALSE, fig.cap = "All Individuals PC1 vs. PC2"}

#extract PCs for individuals
PC_ind <- PC.ind(pca)
eig <- eigenvalues(pca)

#associate metadata with individuals 
PC_ind <- PC_ind %>% rename(Sample=Sample_ID) %>%
  left_join(.,sample_data,by="Sample")

pctot12 <- ggplot(PC_ind, aes(x = Axis1, y = Axis2, color = k_group)) +
    #geom_text(size=3)+
  geom_point(alpha = 0.75, size = 2) +
  labs(x = paste("PC1:", round(eig[1, 3], digits = 3), "%"), 
       y = paste("PC2:", round(eig[2, 3], digits = 3), "%"))+
  scale_color_hue(direction = -1)+
  ggtitle("All_Samples_Axis2vs.Axis1")
pctot12
```

Re-run PCA without k_group 1 (very different from rest)

```{r}
#remove k_group 1
gen2 <- gen1[!gen1@strata$k_group %in% 1]
```

```{r, message=FALSE, warning=FALSE}
#Check PCA eigenvalues
x<- scaleGen(gen2, NA.method = "mean")

# Plot a PCA 
pca <- dudi.pca(x,cent=FALSE,scale=FALSE,scannf=FALSE,nf=4)

#plot1
barplot(pca$eig[1:50],main="All_Samples_PCA_eigenvalues", col=heat.colors(50))
```

```{r message = FALSE,warning = FALSE, fig.cap = "All Individuals PC1 vs. PC2"}

#extract PCs for individuals
PC_ind <- PC.ind(pca)
eig <- eigenvalues(pca)

#associate metadata with individuals 
PC_ind <- PC_ind %>% rename(Sample=Sample_ID) %>%
  left_join(.,sample_data,by="Sample")

pctot12 <- ggplot(PC_ind, aes(x = Axis1, y = Axis2, color = k_group)) +
    #geom_text(size=3)+
  geom_point(alpha = 0.75, size = 2) +
  labs(x = paste("PC1:", round(eig[1, 3], digits = 3), "%"), 
       y = paste("PC2:", round(eig[2, 3], digits = 3), "%"))+
  scale_color_hue(direction = -1)+
  ggtitle("All_Samples_Axis2vs.Axis1")
pctot12
```

Looks like at least three groups in the PCA above

Worth evaluating the PCA if choosing four groups 

## Kmeans and PCA with four groups 

```{r}
#choose four clusters
grp <- find.clusters(gen1, n.pca = 500, n.clust = 4)
grp$size
```

Use the k-means clustering and the sample metadata to create a tibble that contains the 380 individuals that are wild collected and 54 columns of descriptive information, including the kmeans group assignment "k_group"

```{r}
# Returns data frame of individuals and their k-means assignment
x <- as_tibble(grp$grp,rownames=NA) %>%
  rownames_to_column(.,var="Sample") %>%
  rename(k_group = value)
```

```{r}
# add a column of k_means group to data on individuals
sample_data <- sample_info_wild %>%
  left_join(., x, 
              by = "Sample")
```

## PCA - Four K groups 

Use PCAs to confirm there are 4 groups 

```{r}
#identify each individual to each k group in the genind
kg<-sample_data %>%
  select(Sample,k_group)

# Assign strata to genind
strata(gen1) <- kg[match(indNames(gen1),sample_data$Sample),]

# set pop to GG/STX
setPop(gen1) <- ~k_group
```

```{r, message=FALSE, warning=FALSE}
#Check PCA eigenvalues
x<- scaleGen(gen1, NA.method = "mean")

# Plot a PCA 
pca <- dudi.pca(x,cent=FALSE,scale=FALSE,scannf=FALSE,nf=4)

#plot1
barplot(pca$eig[1:50],main="All_Samples_PCA_eigenvalues", col=heat.colors(50))
```

```{r message = FALSE,warning = FALSE, fig.cap = "All Individuals PC1 vs. PC2"}

#extract PCs for individuals
PC_ind <- PC.ind(pca)
eig <- eigenvalues(pca)

#associate metadata with individuals 
PC_ind <- PC_ind %>% rename(Sample=Sample_ID) %>%
  left_join(.,sample_data,by="Sample")

pctot12 <- ggplot(PC_ind, aes(x = Axis1, y = Axis2, color = k_group)) +
    #geom_text(size=3)+
  geom_point(alpha = 0.75, size = 2) +
  labs(x = paste("PC1:", round(eig[1, 3], digits = 3), "%"), 
       y = paste("PC2:", round(eig[2, 3], digits = 3), "%"))+
  scale_color_hue(direction = -1)+
  ggtitle("All_Samples_Axis2vs.Axis1")
pctot12
```

I would say there are four groups based on the K means + PCA.  Proceed with these four. 

## Assign Genetic Groups

First, investigate each k_group and give it a common name based on geographic origin of the majority of the individuals 

```{r}
#K_group1
sample_data %>%
  filter(k_group == 1) %>%
count(SampleOrigin,Geographic_state,Geographic_location, Geographic_code)
```

```{r}

MainStates <- map_data("state")

ggplot()+
  geom_polygon(data=MainStates,aes(x=long,y=lat,group=group),fill="lightgray", color="black") + 
  theme_bw()+
  theme(panel.grid.major = element_blank())+
  theme(panel.grid.minor = element_blank())+
  coord_cartesian(xlim=c(-99.830000,-79.783711), ylim = c(24.783218,30.858945)) + 
 geom_point(data = sample_data %>%
  mutate(latitude2 = round(latitude,3)) %>%
  mutate(longitude2 = round(longitude,3)) %>%
  count(k_group,Geographic_code,longitude2,latitude2) %>%
    mutate_at(c('longitude2', 'latitude2'), as.numeric) %>%
    filter(k_group == 1),
    #counts for proportional representation
             aes(y = latitude2, x = longitude2, group=Geographic_code,size = n), colour= "forestgreen",
             stroke=0.5)
```
k_group 1 is south Texas


```{r}
#k group 2
sample_data %>%
  filter(k_group == 2) %>%
  count(SampleOrigin,Geographic_state,Geographic_location, Geographic_code)
```

```{r}
ggplot()+
  geom_polygon(data=MainStates,aes(x=long,y=lat,group=group),fill="lightgray", color="black") + 
  theme_bw()+
  theme(panel.grid.major = element_blank())+
  theme(panel.grid.minor = element_blank())+
  coord_cartesian(xlim=c(-99.830000,-79.783711), ylim = c(24.783218,30.858945)) + 
 geom_point(data = sample_data %>%
  mutate(latitude2 = round(latitude,3)) %>%
  mutate(longitude2 = round(longitude,3)) %>%
  count(k_group,Geographic_code,longitude2,latitude2) %>%
    mutate_at(c('longitude2', 'latitude2'), as.numeric) %>%
    filter(k_group == 2),
    #counts for proportional representation
             aes(y = latitude2, x = longitude2, group=Geographic_code,size = n), colour= "forestgreen",
             stroke=0.5)
```
k_group 2 is north florida (nfl)

```{r}
sample_data %>%
  filter(k_group == 3) %>%
  count(SampleOrigin,Geographic_state,Geographic_location,Geographic_code)
```

```{r}
ggplot()+
  geom_polygon(data=MainStates,aes(x=long,y=lat,group=group),fill="lightgray", color="black") + 
  theme_bw()+
  theme(panel.grid.major = element_blank())+
  theme(panel.grid.minor = element_blank())+
  coord_cartesian(xlim=c(-99.830000,-79.783711), ylim = c(24.783218,30.858945)) + 
 geom_point(data = sample_data %>%
  mutate(latitude2 = round(latitude,3)) %>%
  mutate(longitude2 = round(longitude,3)) %>%
  count(k_group,Geographic_code,longitude2,latitude2) %>%
    mutate_at(c('longitude2', 'latitude2'), as.numeric) %>%
    filter(k_group == 3),
    #counts for proportional representation
             aes(y = latitude2, x = longitude2, group=Geographic_code,size = n), colour= "forestgreen",
             stroke=0.5)
```
k_group 3 is general florida (gfl)


```{r}
sample_data %>%
  filter(k_group == 4) %>%
  count(SampleOrigin,Geographic_state,Geographic_location,Geographic_code)

```

```{r}
ggplot()+
  geom_polygon(data=MainStates,aes(x=long,y=lat,group=group),fill="lightgray", color="black") + 
  theme_bw()+
  theme(panel.grid.major = element_blank())+
  theme(panel.grid.minor = element_blank())+
  coord_cartesian(xlim=c(-99.830000,-79.783711), ylim = c(24.783218,30.858945)) + 
 geom_point(data = sample_data %>%
  mutate(latitude2 = round(latitude,3)) %>%
  mutate(longitude2 = round(longitude,3)) %>%
  count(k_group,Geographic_code,longitude2,latitude2) %>%
    mutate_at(c('longitude2', 'latitude2'), as.numeric) %>%
    filter(k_group == 4),
    #counts for proportional representation
             aes(y = latitude2, x = longitude2, group=Geographic_code,size = n), colour= "forestgreen",
             stroke=0.5)
```
k_group 4 is northern gulf of mexico (ngom)


Add column in data file that specifies each "k_group" with "genetic group"

```{r}
sample_data <- sample_data %>%
mutate(gen_group = case_when(
    k_group == 1 ~ "stx",
    k_group == 2 ~ "nfl",
    k_group == 3 ~ "gfl",
    k_group == 4 ~ "ngom"))
```

## Summary

The script above takes all of the individuals in the array data and assigns them to genetic groups via kmeans clustering using 20K randomly selected snps.
The data is in the "sample_data" tibble.  The genetic groups are defined as below:

stx = south texas
ngom = northern gulf of mexico
nfl = northern florida
sfl = southern florida

The table below ("summary") summarizes the data

```{r}
summary<- sample_data %>%
  select(Sample,sample_id,Geographic_code,Geographic_state,gen_group) %>%
  group_by(Geographic_state,gen_group) %>%
  count() %>%
  rename(sample_location = Geographic_state, genetic_group = gen_group)
```


## Save files

```{r}
#save file of individual data without selected individuals and with genetic groups assigned through k_means
saveRDS(sample_data, file = "data//derived//sample_data")

#save file of SNP data without selected individuals
saveRDS(genotypes, file = "data//derived//genotypes")

saveRDS(combined_filtered_20K, file = "data//derived//combined_filtered_20K")
```
