# Summary Plots of SNPs for Panel

```{r,include = FALSE}
#library(SNPolisher)
library(tidyverse)
library(hierfstat)
library(adegenet)
library(stringr)
library(ggrepel)
library(pantomime)
library(pegas)
library(patchwork)
library("lattice")
library(maps)
library(mapdata)
library(stringr)
library(data.table)
library(here)

options(scipen = 999)
```

```{r, include = FALSE}
#load in functions for PCA and manipulating genind objects
source("~/PCA.R")
source("~/genind.R")
```

## Distribution of SNPs

```{r}
panel_snps <- readRDS(file = here::here("data","derived","panel_snps"))
as_tibble(panel_snps)
```

```{r}
ggplot(panel_snps, aes(x = chrom, y = pos)) +
  geom_point(size = 0.2)
```

Add in the maximum position for each chromosome (in red)

```{r}
chrom_info <- readRDS(file = here("data","derived","chrom_info"))

p1<-ggplot(panel_snps, aes(x = chrom, y = pos)) +
  geom_point(size = 0.2) +
  geom_point(data = chrom_info, color = "red", size = 1) + 
  xlab("chromosome") + 
  ylab("position") + 
  scale_x_continuous(breaks=1:10) + 
  ggtitle("positions of selected snps per chromosome\nmaximum position in red")
p1
```

```{r}
ggsave(here::here("figures_and_tables","01_snps_along_genome.png"), p1, width = 8, height = 6, units = "in")
```


## MAF per population

Examine the distribution of MAF of SNPs for each population

First fetch the data

```{r}
nfl_frq <- readRDS(file = here::here("data","derived","nfl_frq")) %>%
  rename(snp_id = nfl_probe) %>%
  filter(snp_id %in% panel_snps$snp_id) %>%
  remove_rownames()

gfl_frq <- readRDS(file = here::here("data","derived","gfl_frq")) %>%
  rename(snp_id = gfl_probe) %>%
  filter(snp_id %in% panel_snps$snp_id) %>%
  remove_rownames()

ngom_frq <- readRDS(file = here::here("data","derived","ngom_frq")) %>%
  rename(snp_id = ngom_probe) %>%
  filter(snp_id %in% panel_snps$snp_id) %>%
  remove_rownames()

stx_frq <- readRDS(file = here::here("data","derived","stx_frq")) %>%
  rename(snp_id = stx_probe) %>%
  filter(snp_id %in% panel_snps$snp_id) %>%
  remove_rownames()
```

Format the data

```{r}
maf <- nfl_frq %>%
  left_join(.,gfl_frq,by="snp_id") %>%
  left_join(.,ngom_frq,by="snp_id") %>%
  left_join(.,stx_frq,by="snp_id") %>%
  select(snp_id,everything())
```

Need to format data to "long" format 

```{r}
maf_long <- maf %>%
  rename(stx = stx_frq, ngom = ngom_frq, nfl = nfl_frq, gfl = gfl_frq) %>%
  pivot_longer(!snp_id, names_to = "pop", values_to = "maf") 
```

Plot maf data

```{r}
p2<-ggplot(maf_long, aes(x=pop, y=maf)) + 
  geom_boxplot(width=0.5)+
  xlab("genetic group") + 
  ylab("minor allele frequency") + 
  ggtitle("minor allele frequency of selected snps per genetic group")
p2
```

```{r}
ggsave(here::here("figures_and_tables","02_maf_genetic_groups.png"), p2, width = 8, height = 4, units = "in")
```

## PCA with selected SNPs

Read in sample data

```{r}
sample_data <-readRDS(here("data","derived","sample_data"))
```

Read in genotype data (genotypes for each individual at each loci), filter so only includes loci on panel

```{r}
genotypes <- readRDS(here("data","derived","genotypes_f_he_hw"))
genotypes_with_panel_snps <- genotypes[, maf$snp_id]
```

Turn data into a genind object 

```{r}
gind <- df2genind(genotypes_with_panel_snps, ncode = 1, sep = "", NA.char = NA, ploidy = 2)
gind
``` 

Run a PCA

```{r, message=FALSE, warning=FALSE, fig.show='hide'}
x1 <- scaleGen(gind, NA.method = "mean")

pca1 <- dudi.pca(x1,cent=FALSE,scale=FALSE,scannf=FALSE,nf=4)

eig1<-barplot(pca1$eig[1:10],main="PCA eigenvalues", col=heat.colors(50))
```

```{r,message=FALSE}
PC_ind <- PC.ind(pca1)
eig <- eigenvalues(pca1)

#rename "Sample_ID" in PC_ind to "INDV"
PC_ind <- PC_ind %>% rename(Sample=Sample_ID)

PC_ind <- merge(PC_ind,sample_data,by = "Sample") 

pc <- ggplot(PC_ind, aes(x = Axis1, y = Axis2, color = gen_group)) +
  geom_point(alpha = 0.75, size = 2) +
  labs(x = paste("PC1:", round(eig[1, 3], digits = 3), "%"), 
       y = paste("PC2:", round(eig[2, 3], digits = 3), "%"))+
  ggtitle("5,250 candidate snps and gulf individuals from 600K array") +
  scale_color_hue(direction = -1)
pc
```


```{r}
ggsave(here::here("figures_and_tables","03_pca_gulf_from_600K_array.png"), pc, width = 8, height = 6, units = "in")
```
